Think deep for reef resilience

### 2025-08-14 paper/paper_draft.qmd
Update DHW limitations section
Add comment for @ConnectedSystems

Our findings suggest that reefs in ‘high’ cover clusters do not have greater thermal
tolerance than reefs in lower coral cover clusters. This is in contrast with literature
suggesting corals have the capacity to adapt to increasing climate change pressures
[@humanes2022; @lachs2024], including the investigation by Bairos-Novak et al.
[-@bairos-novak2021] that revealed high heritability of bleaching and survival response
traits, used to inform thermal tolerance trait distributions in this investigation. The low
importance of thermal tolerance on cover outcomes may be due to reefs experiencing high
levels of thermal stress early in projected timeseries. High - early timeseries thermal
stress events may prevent the intergenerational adaptation of corals to thermal stress and
prevent thermal tolerance from reaching levels greater than 9°C-weeks. **\[TI note: This
paragraph needs to be revisited\]**

@@

There are multiple caveats in how coral ecosystem dynamics are modelled in this
investigation. We model coral colonies as circular diameter values that are tracked over
timesteps through growth. Due to the availability of quality future projected data we do not
consider the impacts of waves, cyclones, currents and Crown of Thorns Starfish (CoTS) on
coral colony survival and bleaching [@massel1993; @deaker2022]. Uncertainty in connectivity
modelling outputs, satellite-derived bathymetry data and future thermal stress projections
are likely to influence the results obtained in this investigation. While the resolution of
connectivity data used in this investigation improves upon data used in comparable studies
[@bozec2025], the time-mean implementation used may not effectively capture long term
patterns [@vogt-vincent2023]. Using alternative connectivity implementation methods that
better capture the ecological-timescale trends in reef connections rather than the high
daily variability, would improve confidence in the connectivity findings of this
investigation [@vogt-vincent2023]. The DHW data used in this investigation have not been
bias corrected for the Great Barrier Reef region and are part of a wider Australian dataset
[@dixon2022] developed using statistical downscaling. We investigate a range of future
climate responses, thermal projections and ecological parameterisations of ADRIA - CoralBlox
to account for many possible futures, and our results suggest the factors that influence
reef resilience are consistent across these scenarios.

@@

**Bleaching mortality**

In ADRIA we track the thermal tolerance of each functional group and size class as a
truncated normal distribution that is updated in each timestep following growth and
recruitment. This distribution determines the proportion of the population on a reef that is
susceptible to bleaching (i.e. has a thermal threshold below the experienced DHW value).
Starting distributions for thermal tolerance levels are informed by Bairos-Novak et al.
[-@bairos-novak2021].

>>>>

Our findings suggest that reefs in ‘high’ cover clusters do not have greater thermal
tolerance than reefs in lower coral cover clusters. This is in contrast with literature
suggesting corals have the capacity to adapt to increasing climate change pressures
[@humanes2022; @lachs2024; @lachs2023], including the investigation by Bairos-Novak et al.
[-@bairos-novak2021] that revealed high heritability of bleaching and survival response
traits, used to inform thermal tolerance trait distributions in this investigation. The low
importance of thermal tolerance on cover outcomes may be due to reefs experiencing high
levels of thermal stress early in projected timeseries. High - early timeseries thermal
stress events may prevent the intergenerational adaptation of corals to thermal stress and
prevent thermal tolerance from reaching levels greater than 9°C-weeks. **\[TI note: This
paragraph needs to be revisited\]**

@@

There are multiple caveats in how coral ecosystem dynamics are modelled in this
investigation. We model coral colonies as circular diameter values that are tracked over
timesteps through growth. Due to the availability of quality future projected data we do not
consider the impacts of waves, cyclones, currents and Crown of Thorns Starfish (CoTS) on
coral colony survival and bleaching [@massel1993; @deaker2022]. Uncertainty in connectivity
modelling outputs, satellite-derived bathymetry data and future thermal stress projections
are likely to influence the results obtained in this investigation. While the resolution of
connectivity data used in this investigation improves upon data used in comparable studies
[@bozec2025], the time-mean implementation used may not effectively capture long term
patterns [@vogt-vincent2023]. Using alternative connectivity implementation methods that
better capture the ecological-timescale trends in reef connections rather than the high
daily variability, would improve confidence in the connectivity findings of this
investigation [@vogt-vincent2023]. The DHW data used in this investigation have not been
bias corrected for the Great Barrier Reef region and are part of a wider Australian dataset
[@dixon2022] developed using statistical downscaling. Additionally, Lachs et al.
[@lachs2021] suggest the utility of DHW levels for prediction of coral bleaching can be
improved by reducing the threshold for HotSpot Anomaly values from MMM +1 to MMM, and
reducing the accumulation window from 12 weeks to 8 weeks. We investigate a range of future
climate responses, thermal projections and ecological parameterisations of ADRIA - CoralBlox
to account for many possible futures, and our results suggest the factors that influence
reef resilience are consistent across these scenarios.

@@

**Bleaching mortality**
**\[BG: Has this section been completed?\]**
In ADRIA we track the thermal tolerance of each functional group and size class as a
truncated normal distribution that is updated in each timestep following growth and
recruitment. This distribution determines the proportion of the population on a reef that is
susceptible to bleaching (i.e. has a thermal threshold below the experienced DHW value).
Starting distributions for thermal tolerance levels are informed by Bairos-Novak et al.
[-@bairos-novak2021].
